A general and efficient representation of Ancestral Recombination Graphs

### 2023-10-06 paper.tex
Minor updates

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A classical event ARG. (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$.
(C) A concrete encoding of the information (right).
Recombination nodes have a non-null breakpoint
and two parents.
}
\end{figure}

@@

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks.
The vitally important and fundamental distinction between
a gARG and an eARG is that gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, which cannot be done
by specifying recombination events and breakpoints.
% FIXME this sentence is problematic. We should recast to talk about
% the outcomes of events, not talk about simulateousness (which will
% confuse people
Thus, we have the means of encoding the effects of
multiple \emph{simultaneous} events, and to
systematically describe the uncertainty around their
ordering (see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

@@

\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward
(Fig.~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and then add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the outbound edges with the intervals $[0, x)$
and $[x, L)$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

@@

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
independently and subsequently stitching them together
into an ARG.
Two recent methods take this approach,
\relate\ and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and then identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among small samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

@@

\subsection*{The Big and Little ARG}
\label{app-big-and-little-arg}

@@

\subsection*{Survey of ARG inference methods}

>>>>

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A classical event ARG. (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see section XXX for a discussion
of this important point).
(C) A concrete encoding of the information (right).
Recombination nodes have a non-null breakpoint
and two parents.
}
\end{figure}

@@

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks.
The vitally important and fundamental distinction between
a gARG and an eARG is that gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, which cannot be done
by specifying recombination events and breakpoints.
% FIXME this sentence is problematic. We should recast to talk about
% the outcomes of events, not talk about simulateousness (which will
% confuse people
Thus, we have the means of encoding the \emph{effects} of
multiple evolutionary events, and to
systematically describe the uncertainty around their
ordering (see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

@@

\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward (Fig.~\ref{fig-ancestry-resolution}B):
we duplicate the nodes and
edges from the input eARG, and then add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the two outbound edges with the intervals $[0, x)$
and $[x, L)$, respectively.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

@@

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
independently and subsequently stitching them together
into an ARG.
Two recent methods take this approach,
\relate~\citep{speidel2019method} and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and then identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among small samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

@@

\section{The Big and Little ARG}
\label{app-big-and-little-arg}

@@

\section{Survey of ARG inference methods}
